A generalization of the Bradley-Terry model for draws in chess

### 2019-05-13 inst/collusion_chess_draws.Rnw
more work, including correcting a transpose error on the third subtable of the kka results

\cite{moul2009}: for instance, a player may have the choice of more
and less conservative strategies with different chances for a
draw. Risk- averse players may pursue less aggressive strategies,
trading expected payout for reduced variance

@@

In inference problems where the dataset is outcomes of paired
comparisons, such as chess, the Bradley-Terry model can be augmented
to accommodate draws in a number of ways.  Here I present a new
augmentation of Bradley-Terry in which a drawn chess match is regarded
as a three-way competition between the players and an entity that wins
if the match is drawn.  The resulting likelihood functions are easily
optimised numerically and I present a number of results.

@@

\begin{equation}
\prod_{i<j}
\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}^\mathrm{white}}
\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}^\mathrm{black}}
\end{equation}

\begin{equation}
\prod_{i<j}
\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}^\mathrm{white}+D_{ij}^\mathrm{black}}
\end{equation}

which might be preferable on grounds of clarity.  We now apply this
likelihood model to real chess datasets.  The general approach to
testing any hypothesis $H$ on any likelihood function $\mathcal{L}$ is
first to identify the constraint on hypothesis space corresponding to
$H$.  One then finds the unconstrained maximum likelihood
$\mathcal{L}_\mathrm{free}$ and compares it with the maximum
likelihood $\mathcal{L}_\mathrm{cons}$ when the optimization is
constrained by $H$.  The test statistic is then the likelihood ratio
$\mathcal{L}_\mathrm{free}/\mathcal{L}_\mathrm{cons}$ or, equivalently
the support $\mathcal{S}=\log\mathcal{R}$.  Wilks's theorem states
that the asymptotic distribution of $2\mathcal{S}$ is $\chi^2_n$ where
$n$ is the number of degrees of freedom implemented by $H$.

@@

\begin{table}
\centering
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays White,}\\ \small{wins}} & \multicolumn{3}{c@{}}{\small{Plays black, loses}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &  18 &   6  \\
  Kr &  7 &  -  &  18  \\
  Ks & 15 &  30 &  -   \\
\end{tabular}\quad
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays W,}\\ \small{draws}} & \multicolumn{3}{c@{}}{\small{Plays black, draws}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &  20 &  20  \\
  Kr & 29 &  -  &  72  \\
  Ks & 26 &  57 &  -   \\
\end{tabular}\quad
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays W,}\\ \small{loses}} & \multicolumn{3}{c@{}}{\small{Plays black, wins}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &  13 &   2  \\
  Kr &  5 &  -  &   7  \\
  Ks & 11 &   9 &  -   \\
\end{tabular}
\caption{Results of 365 chess matches \label{kka} between Grandmaster
  Anand, Grandmaster Karpov, and Grandmaster Kasparov.  Thus, for
  example, Anand played white against Karpov a total of $18+20+13=51$
  times and won 18, drew 20, lost 13}
\end{table}

@@

\begin{equation}
\mathcal{L}\left(
p_\mathrm{An},p_\mathrm{Kr},p_\mathrm{Ks}
,W,
d_\mathrm{An},d_\mathrm{Kr},d_\mathrm{Ks}
\right)=
\frac{
\parbox{3in}{$
p_\mathrm{An}^{15}
p_\mathrm{Kr}^{12}
p_\mathrm{Ks}^{20}
\left(p_\mathrm{Kr} + W\right)^{25}
\left(p_\mathrm{Ks} + W\right)^{45} 
\left(p_\mathrm{An} + W\right)^{24}\\ \rule{10mm}{0mm}
\left(d_\mathrm{Kr} + d_\mathrm{Ks}\right)^{129}
\left(d_\mathrm{Kr} + d_\mathrm{An}\right)^{49}
\left(d_\mathrm{Ks} + d_\mathrm{An}\right)^{46}
$}}{\parbox{3in}{$
\left(p_\mathrm{Kr} + p_\mathrm{Ks} + W + d_\mathrm{Kr} + d_\mathrm{Ks}\right)^{193}\\ \rule{10mm}{0mm}
\left(p_\mathrm{Kr} + p_\mathrm{An} + W + d_\mathrm{Kr} + d_\mathrm{An}\right)^{92}\\ \rule{20mm}{0mm}
\left(p_\mathrm{Ks} + p_\mathrm{An} + W + d_\mathrm{Ks} + d_\mathrm{An}\right)^{80}
$}}
\end{equation}

@@

We see considerable variability in proclivity to draw and again this
may be tested with the Method of Support, which gives a $p$-value of
about 0.039, borderline significant: an interesting result, given that
in this case no
player's draw monster appears alone anywhere in the likelihood function.

@@

\section{Chess collusion: Cura\c{c}ao 1962}
\begin{table}
\centering
\begin{tabular}{ll|cccccccc}
 &
&\rotatebox{-90}{Petrosian} 
&\rotatebox{-90}{Keres} 
&\rotatebox{-90}{Geller}
&\rotatebox{-90}{Fischer} 
&\rotatebox{-90}{Korchnoi}
&\rotatebox{-90}{Benko}
&\rotatebox{-90}{Tal}
&\rotatebox{-90}{Filip}\\
Petrosian &USSR    &             &DDDD    &DDDD   &DWDD   &DDWW   &DDWD   &WWD    &DWWD\\
Keres     &USSR    &             &NA      &DDDD   &LDWD   &DDWD   &WWWL   &WDW    &DWWD\\
Geller    &USSR    &             &NA      &NA     &WWDL   &DDWD   &DDDW   &DWW    &DWWD\\
Fischer   &USA     &             &NA      &NA     &NA     &LWLD   &LWDW   &DWD    &WDWD\\
Korchnoi  &USSR    &             &NA      &NA     &NA     &NA     &DDDL   &WLD    &WWWW\\
Benko     &USA     &             &NA      &NA     &NA     &NA     &NA     &WLD    &LWWD\\
Tal       &USSR    &             &NA      &NA     &NA     &NA     &NA     &NA     &WLD \\
Filip     &TCH     &             &NA      &NA     &NA     &NA     &NA     &NA     &A \\
\end{tabular}
\caption{Curacao 1962 SPELLED WRONG}
\end{table}

@@

>>>>

\cite{moul2009}: for instance, a player may have the choice of more
and less conservative strategies with different chances for a
draw.  Risk- averse players may pursue less aggressive strategies,
trading expected payout for reduced variance.

moul: White's relative advantage over Black will be dependent upon the
comparative mastery of the two players.

The most obvious issue regarding the threshold against which we compare
this relative advantage of White is the inherent advantage of White's
position in moving first. We correspondingly do not expect these base
thresholds to be symmetric and impose no such restriction.

@@

In inference problems where the dataset is outcomes of paired
comparisons, such as chess, the Bradley-Terry
model~\citep{bradley1952} can be augmented to accommodate draws in a
number of ways.  Here I present a new augmentation of Bradley-Terry in
which a drawn chess match is regarded as a three-way competition
between the players and an entity that wins if the match is drawn.
The resulting likelihood functions are easily optimised numerically
and I present a number of results.

@@

\begin{equation}
\prod_{i<j}
\left(\frac{p_D}{p_i+p_j+p_D+p_W}\right)^{D_{ij}^\mathrm{white}}
\left(\frac{p_D}{p_i+p_j+p_D+p_W}\right)^{D_{ij}^\mathrm{black}}
=
\prod_{i<j}
\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}^\mathrm{white}+D_{ij}^\mathrm{black}}
\end{equation}

which might be preferable on grounds of clarity.  We now apply this
and related likelihood models to real chess datasets.  The general
approach to testing any hypothesis $H$ on any likelihood function
$\mathcal{L}$ is first to identify the constraint on hypothesis space
corresponding to $H$.  One then finds the unconstrained maximum
likelihood $\mathcal{L}_\mathrm{free}$ and compares it with the
maximum likelihood $\mathcal{L}_\mathrm{cons}$ when the optimization
is constrained by $H$.  The test statistic is then the likelihood
ratio $\mathcal{L}_\mathrm{free}/\mathcal{L}_\mathrm{cons}$ or,
equivalently the support $\mathcal{S}=\log\mathcal{R}$.  Wilks's
theorem states that the asymptotic distribution of $2\mathcal{S}$ is
$\chi^2_n$ where $n$ is the number of degrees of freedom implemented
by $H$.

\subsection{Discussion}

The above likelihood functions are objective and reasonable
representations of chess match results.  However, they differ markedly
from previous likelihood functions of draws in that they include
entities representing white player advantage ($p_W$), and the
proclivity to draw $p_D$).   The $p_W$

@@

\begin{table}
\centering
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays White,}\\ \small{wins}} & \multicolumn{3}{c@{}}{\small{Plays black, loses}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &  18 &   6  \\
  Kr &  7 &  -  &  18  \\
  Ks & 15 &  30 &  -   \\
\end{tabular}\quad
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays W,}\\ \small{draws}} & \multicolumn{3}{c@{}}{\small{Plays black, draws}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &  20 &  20  \\
  Kr & 29 &  -  &  72  \\
  Ks & 26 &  57 &  -   \\
\end{tabular}\quad
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays W,}\\ \small{loses}} & \multicolumn{3}{c@{}}{\small{Plays black, wins}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &   5 &   11  \\
  Kr & 13 &  -  &    9  \\
  Ks &  2 &   7 &  -   \\
\end{tabular}
\caption{Results of 365 chess matches \label{kka} between Grandmaster
  Anand, Grandmaster Karpov, and Grandmaster Kasparov.  Thus, for
  example, Anand played white against Karpov a total of $18+20+13=51$
  times and won 18, drew 20, lost 13}
\end{table}

@@

\begin{equation}\label{KKA_personalised_drawmonster}
\mathcal{L}\left(
p_\mathrm{An},p_\mathrm{Kr},p_\mathrm{Ks}
,W,
d_\mathrm{An},d_\mathrm{Kr},d_\mathrm{Ks}
\right)=
\frac{
\parbox{3in}{$
p_\mathrm{An}^{15}
p_\mathrm{Kr}^{12}
p_\mathrm{Ks}^{20}
\left(p_\mathrm{Kr} + W\right)^{25}
\left(p_\mathrm{Ks} + W\right)^{45} 
\left(p_\mathrm{An} + W\right)^{24}\\ \rule{10mm}{0mm}
\left(d_\mathrm{Kr} + d_\mathrm{Ks}\right)^{129}
\left(d_\mathrm{Kr} + d_\mathrm{An}\right)^{49}
\left(d_\mathrm{Ks} + d_\mathrm{An}\right)^{46}
$}}{\parbox{3in}{$
\left(p_\mathrm{Kr} + p_\mathrm{Ks} + W + d_\mathrm{Kr} + d_\mathrm{Ks}\right)^{193}\\ \rule{10mm}{0mm}
\left(p_\mathrm{Kr} + p_\mathrm{An} + W + d_\mathrm{Kr} + d_\mathrm{An}\right)^{92}\\ \rule{20mm}{0mm}
\left(p_\mathrm{Ks} + p_\mathrm{An} + W + d_\mathrm{Ks} + d_\mathrm{An}\right)^{80}
$}}
\end{equation}

@@

We see considerable variability in proclivity to draw and again this
may be tested with the Method of Support, which gives a $p$-value of
about 0.039, borderline significant: an interesting result, given that
in this case no draw monster appears alone anywhere in the likelihood
function.

@@

\subsection{Each player has a distinct White monster}

We now consider the possibility that the strength conferred by playing
White, is specific to each player and we denote this $w_\mathrm{An}$,
$w_\mathrm{Kr}$ and $w_\mathrm{Ks}$.  The corresponding likelihood
function would be

\begin{equation}\label{KKA_personalised_whitemonster}
%\mathcal{L}\left(
%p_\mathrm{An},p_\mathrm{Kr},p_\mathrm{Ks},
%w_\mathrm{An},w_\mathrm{Kr},w_\mathrm{Ks},D
%\right)=
\frac{
%\parbox{3in}
%       {$
p_\mathrm{An}^{15}
p_\mathrm{Kr}^{12}
p_\mathrm{Ks}^{20}
D^{224}
\left(p_\mathrm{An} + w_\mathrm{An}\right)^{24}%\\ \rule{10mm}{1mm}
\left(p_\mathrm{Kr} + w_\mathrm{Kr}\right)^{25}
\left(p_\mathrm{Ks} + w_\mathrm{Ks}\right)^{45}
%$}
}{\parbox{4in}{$
\left(p_\mathrm{An} + p_\mathrm{Kr} + w_\mathrm{An} +  D\right)^{43}
\left(p_\mathrm{An} + p_\mathrm{Kr} + w_\mathrm{Kr} +  D\right)^{49}\\ \rule{10mm}{0mm}
\left(p_\mathrm{An} + p_\mathrm{Ks} + w_\mathrm{An} +  D\right)^{37}
\left(p_\mathrm{An} + p_\mathrm{Ks} + w_\mathrm{Ks} +  D\right)^{43}\\ \rule{20mm}{0mm}
\left(p_\mathrm{Kr} + p_\mathrm{Ks} + w_\mathrm{Kr} +  D\right)^{99}
\left(p_\mathrm{Kr} + p_\mathrm{Ks} + w_\mathrm{Ks} +  D\right)^{94}
$}}
\end{equation}

In the likelihood function above we see, for example, that Anand won a
total of $18+6=24$ matches playing White, $13+2=15$ games playing
Black, and that Anand played white against Karpov a total of
$18+20+5=43$ times.  We fail to reject the null of
$w_\mathrm{An}=w_\mathrm{Kr}=w_\mathrm{Ks}$, having a likelihood ratio
of $4.5$, or an asymptotic $p$-value of $0.076$.  There is no strong
evidence that the players differ in regard to their white advantage.

\section{Chess collusion: Cura\c{c}ao 1962}
\begin{table}
\centering
\begin{tabular}{ll|cccccccc}
 &
&\rotatebox{-90}{Petrosian} 
&\rotatebox{-90}{Keres} 
&\rotatebox{-90}{Geller}
&\rotatebox{-90}{Fischer} 
&\rotatebox{-90}{Korchnoi}
&\rotatebox{-90}{Benko}
&\rotatebox{-90}{Tal}
&\rotatebox{-90}{Filip}\\
Petrosian &USSR    &             &DDDD    &DDDD   &DWDD   &DDWW   &DDWD   &WWD    &DWWD\\
Keres     &USSR    &             &NA      &DDDD   &LDWD   &DDWD   &WWWL   &WDW    &DWWD\\
Geller    &USSR    &             &NA      &NA     &WWDL   &DDWD   &DDDW   &DWW    &DWWD\\
Fischer   &USA     &             &NA      &NA     &NA     &LWLD   &LWDW   &DWD    &WDWD\\
Korchnoi  &USSR    &             &NA      &NA     &NA     &NA     &DDDL   &WLD    &WWWW\\
Benko     &USA     &             &NA      &NA     &NA     &NA     &NA     &WLD    &LWWD\\
Tal       &USSR    &             &NA      &NA     &NA     &NA     &NA     &NA     &WLD \\
Filip     &TCH     &             &NA      &NA     &NA     &NA     &NA     &NA     &A \\
\end{tabular}
\caption{Cura\c{c}ao 1962 candidates matches, world championship 1963}
\end{table}

@@

\section{Conclusions}

In this context we note, following \cite{langer2013}, that a player
may well adopt a riskier strategy after a sequence of defeats,
adopting a ``nothing-to-lose'' mentality.  Such a phenomenon would be
detectable, at least in principle, by generalizing the personalised
draw likelihood function \ref{KKA_personalised_drawmonster} and
seeking evidence for the draw monster's strength changing as a
tournament progresses: it would become stronger for a player on a
winning streak, and weaker for a losing player.
